---
title: "Lab 3: COVID-19"
subtitle: 'Ecosystem Science and Sustainability 330'
author:
  - name: Mia Colangelo
    email: miacola@colostate.edu
format:
  html:
    self-contained: true
execute:
  echo: true
---

```{r}
library(tidyverse)
library(flextable)
library(zoo)
library(ggplot2)
library(readr)
library(dplyr)
```

```{r}
url <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv"
covid_data <- read.csv(url)
glimpse(covid_data)
```

Question 1: Take a moment to reflect on the value of open data: How does easy access to historical and real-time environmental data shape our understanding of climate trends, resource management, and public health? What happens when this data disappears or becomes inaccessible? The role of independent archiving and collaborative stewardship has never been more critical in ensuring scientific progress and accountability.

*Access to truthful and current data is critical to our understanding of the world around us because it aids us in making connections between global trends. When this data is not accessible and understandable to the public, it breeds ignorance and misinformation. People cannot fact check authority figures or press, and researchers cannot rely on poorly constructed data.*

```{r}
my_date <- as.Date("2022-01-01")
my_state <- "Colorado"
latest_date <- max(covid_data$date, na.rm = TRUE)


## Question 2

colorado_data <- covid_data %>%
  filter(state == my_state) %>%
  arrange(county, date) %>%
  group_by(county) %>%
  mutate(new_cases = cases - lag(cases, default = first(cases)),
         new_deaths = deaths - lag(deaths, default = first(deaths))) %>%
  ungroup()

worst_cumulative <- colorado_data %>%
  filter(date == my_date) %>%
  arrange(desc(cases)) %>%
  head(5)

worst_new <- colorado_data %>%
  filter(date == my_date) %>%
  arrange(desc(new_cases)) %>%
  head(5)

worst_cumulative %>% select(county, cases) %>% flextable() %>% set_caption("Top 5 Counties by Cumulative Cases")
worst_new %>% select(county, new_cases) %>% flextable() %>% set_caption("Top 5 Counties by New Cases")


safe_counties <- colorado_data %>%
  filter(date == latest_date, new_cases == 0) %>%
  select(county)

num_safe_counties <- nrow(safe_counties)
flextable(safe_counties) %>% set_caption("Counties with 0 New Cases")

total_new_cases <- sum(colorado_data$new_cases[colorado_data$date == latest_date], na.rm = TRUE)
total_cumulative_cases <- sum(colorado_data$cases[colorado_data$date == latest_date], na.rm = TRUE)

report_text <- glue::glue("
### COVID-19 Summary Report for {my_state} ({latest_date})

- **Total cumulative cases:** {total_cumulative_cases}
- **Total new cases today:** {total_new_cases}
- **Number of safe counties (0 new cases):** {num_safe_counties}

Stay safe. Monitor updates regularly.
")

# Print report
cat(report_text)


```

Question 2:

```{r}
## Question 3
population_data <- read_csv("/Users/miacolangelo/Desktop/ESS 330/github/lab3/data/co-est2023-alldata.csv")

population_data <- population_data %>%
  mutate(STATE = as.numeric(STATE)) %>%

  mutate(
    STATE = sprintf("%02d", STATE),
    COUNTY = sprintf("%03s", COUNTY),
    FIPS = paste0(STATE, COUNTY)
  ) %>%
  select(FIPS, NAME, starts_with("2021")) %>%
  filter(COUNTY != "000")
  
population_data <- population_data %>%
  mutate(FIPS = paste0(sprintf("%02s", STATE), sprintf("%03s", COUNTY))) %>%
  filter(COUNTY != "000") %>%
  select(FIPS, NAME, starts_with("2021"))



population_data <- population_data %>%
  mutate(STATE = sprintf("%02d", as.numeric(STATE)),
         COUNTY = sprintf("%03d", as.numeric(COUNTY),
         FIPS = paste0(STATE, COUNTY)) %>%
  select(matches("NAME|2021"), FIPS) %>%
  filter(COUNTY != "000")
  
colorado_data <- colorado_data %>%
  left_join(population_data, by = c("fips" = "FIPS")) %>%
  mutate(per_capita_cases = cases / `2021`,
         per_capita_new_cases = new_cases / `2021`,
         per_capita_new_deaths = new_deaths / `2021`)

top_5_cumulative_per_capita <- colorado_data %>%
  filter(date == "2021-01-01") %>%
  arrange(desc(per_capita_cases)) %>%
  select(county, per_capita_cases) %>%
  head(5)

top_5_new_cases_per_capita <- colorado_data %>%
  filter(date == "2021-01-01") %>%
  arrange(desc(per_capita_new_cases)) %>%
  select(county, per_capita_new_cases) %>%
  head(5)

# Count safe counties (where new cases = 0)
num_safe_counties <- colorado_data %>%
  filter(date == max(date, na.rm = TRUE), new_cases == 0) %>%
  nrow()

# Generate Report
report <- paste0(
  "COVID-19 Report for Colorado\n",
  "Total new cases on latest date: ", sum(colorado_data$new_cases, na.rm = TRUE), "\n",
  "Total cumulative cases: ", sum(colorado_data$cases, na.rm = TRUE), "\n",
  "Number of safe counties (no new cases): ", num_safe_counties
)

# Print Report
cat(report)

library(flextable)

# Format tables
flextable(top_5_cumulative) %>% set_caption("Top 5 Counties by Cumulative Cases")
flextable(top_5_new_cases) %>% set_caption("Top 5 Counties by New Cases")

flextable(top_5_cumulative_per_capita) %>% set_caption("Top 5 Counties by Cumulative Cases Per Capita")
flextable(top_5_new_cases_per_capita) %>% set_caption("Top 5 Counties by New Cases Per Capita")

```

Question 3:

```{r}
## Question 4

recent_data <- colorado_data %>%
  filter(date >= max(date) - 14)
rolling_threshold <- recent_data %>%
  group_by(county) %>%
  summarize(
    new_cases_14_days = sum(daily_new_cases, na.rm = TRUE),
    per_100k = (new_cases_14_days / `2021`) * 100000
  ) %>%
  arrange(desc(per_100k))
watchlist_count <- rolling_threshold %>%
  filter(per_100k > 100) %>%
  nrow()

recent_data <- colorado_data %>%
  filter(date >= max(date) - 14)

rolling_threshold <- recent_data %>%
  group_by(county) %>%
  summarize(
    new_cases_14_days = sum(new_cases, na.rm = TRUE),
    per_100k = (new_cases_14_days / `2021`) * 100000
  ) %>%
  arrange(desc(per_100k))

top_5_counties <- rolling_threshold %>% head(5)
print(top_5_counties)

watchlist_count <- rolling_threshold %>%
  filter(per_100k > 100) %>%
  nrow()
cat("Number of counties on the watch list:", watchlist_count)

```

Question 4:

```{r}
## Question 5 

death_analysis <- colorado_data %>%
  filter(format(date, "%Y") == "2021") %>%
  group_by(county) %>%
  summarize(
    total_covid_deaths = sum(daily_new_deaths, na.rm = TRUE),
    total_deaths_2021 = max(`2021_total_deaths`, na.rm = TRUE),
    covid_death_percentage = (total_covid_deaths / total_deaths_2021) * 100
  ) %>%
  filter(covid_death_percentage >= 20)
ggplot(death_analysis, aes(x = county, y = covid_death_percentage, fill = county)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(
    title = "Counties Where COVID Deaths Accounted for ≥20% of Total Deaths (2021)",
    x = "County",
    y = "COVID Death Percentage"
  )

```

Question 5:

```{r}
## Question 6

state_data <- covid_data %>%
  group_by(state, date) %>%
  summarize(daily_new_cases = sum(cases - lag(cases, default = 0), na.rm = TRUE)) %>%
  ungroup() %>%
  filter(state %in% c("New York", "Colorado", "Alabama", "Ohio"))

state_data <- state_data %>%
  group_by(state) %>%
  mutate(rolling_avg_cases = rollmean(daily_new_cases, k = 7, fill = NA, align = "right")) %>%
  ungroup()

library(ggplot2)

ggplot(state_data, aes(x = date, y = daily_new_cases)) +
  geom_col(fill = "skyblue") +
  geom_line(aes(y = rolling_avg_cases), color = "red", size = 1) +
  facet_wrap(~ state, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Daily New COVID-19 Cases and 7-Day Rolling Average",
    x = "Date",
    y = "Number of Cases"
  )

state_data_per_capita <- state_data %>%
  left_join(population_data, by = "state") %>%
  mutate(
    cases_per_capita = daily_new_cases / population * 100000,
    rolling_cases_per_capita = rollmean(cases_per_capita, k = 7, fill = NA, align = "right")
  )

ggplot(state_data_per_capita, aes(x = date, y = rolling_cases_per_capita, color = state, group = state)) +
  geom_line(size = 1) +
  theme_minimal() +
  labs(
    title = "7-Day Rolling Average of COVID-19 Cases Per Capita",
    x = "Date",
    y = "Cases Per 100,000 People",
    color = "State"
  )

```

Question 6:

```{r}
## Question 7

readr::read_csv("https://raw.githubusercontent.com/mikejohnson51/csu-ess-330/refs/heads/main/resources/county-centroids.csv")

centroids <- read_csv("https://raw.githubusercontent.com/mikejohnson51/csu-ess-330/refs/heads/main/resources/county-centroids.csv")

covid_with_location <- covid_data %>%
  left_join(centroids, by = "fips")

mean_center <- covid_with_location %>%
  group_by(date) %>%
  summarize(
    weighted_lon = sum(LON * cases, na.rm = TRUE) / sum(cases, na.rm = TRUE),
    weighted_lat = sum(LAT * cases, na.rm = TRUE) / sum(cases, na.rm = TRUE),
    total_cases = sum(cases, na.rm = TRUE),
    month = format(date, "%m")
  )

ggplot(data = mean_center, aes(x = weighted_lon, y = weighted_lat, color = month, size = total_cases)) +
  borders("state", fill = "gray90", colour = "white") +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Weighted Mean Center of COVID-19 Cases in the USA",
    x = "Longitude",
    y = "Latitude",
    color = "Month",
    size = "Total Cases"
  )

ggsave(
  filename = "weighted_mean_center.png", 
  plot = last_plot(),
  width = 10, 
  height = 6, 
  dpi = 300
)


```

Question 7: In a few sentences, describe the movement of the COVID-19 weighted mean throughout the USA and possible drivers of its movement given your knowledge of the outbreak hot spots.

Question 8: As extra credit, extend your analysis in problem three to also compute the weighted mean center of daily COVID deaths. Make two plots next to each other (using patchwork) showing cases in red and deaths in navy. Once completed describe the differences in the plots and what they mean about the spatial patterns seen with COVID impacts.
